```{r setup, include=FALSE}
# set global chunk options
library(knitr)
opts_chunk$set(cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE)
# import functions
source("functions.R")
load("ic_ec.Rdata")
```

Repeating the analysis done in STATA
------------------------------------
### Divide the flu shots in categories, ANOVA for the antibody titers.

#### IC50
We define categories for shots: no shots, between 1 and 3, more than 3.
```{r define_shot_ternary_ic}
ternary_shot <- rep(NA, length(ic_data$flushot_avg))
ternary_shot[ic_data$flushot_avg == 0] <- 'cat0'
ternary_shot[ic_data$flushot_avg >= 1 & ic_data$flushot_avg <= 3] <- 'cat1'
ternary_shot[ic_data$flushot_avg > 3] <- 'cat2'
ic_data$ternary_shot <- factor(ternary_shot)
```

Here below whether different categories have different titers.

```{r shot_categories, echo=TRUE}
fit_H1_B <- aov(H1_brisbane_log ~ ternary_shot, data=ic_data)
summary(fit_H1_B)

fit_H1_P <- aov(H1_PR8_log ~ ternary_shot, data=ic_data)
summary(fit_H1_P)

fit_H2 <- aov(H2_log ~ ternary_shot, data=ic_data)
summary(fit_H2)

fit_H3_B <- aov(H3_brisbane_log ~ ternary_shot, data=ic_data)
summary(fit_H3_B)

fit_H3_H <- aov(H3_HK_log ~ ternary_shot, data=ic_data)
summary(fit_H3_H)

fit_H4 <- aov(H4_log ~ ternary_shot, data=ic_data)
summary(fit_H4)
```

#### EC50
Same as above 
```{r define_shot_ternary_ec}
# Obviously the shots must match
all(ec_data$flushot_avg == ic_data$flushot_avg)
ec_data$ternary_shot <- factor(ternary_shot)
```

```{r shot_categories_ec, echo=TRUE}
fit_H1_pdm <- aov(H1_pdm09_log ~ ternary_shot, data=ec_data)
summary(fit_H1_pdm)

fit_H1_PR8 <- aov(H1_PR8_log ~ ternary_shot, data=ec_data)
summary(fit_H1_PR8)

fit_H12 <- aov(H12_log ~ ternary_shot, data=ec_data)
summary(fit_H12)

fit_H2 <- aov(H2_log ~ ternary_shot, data=ec_data)
summary(fit_H2)

fit_H3 <- aov(H3_log ~ ternary_shot, data=ec_data)
summary(fit_H3)

fit_H4 <- aov(H4_log ~ ternary_shot, data=ec_data)
summary(fit_H4)

fit_H5 <- aov(H5_log ~ ternary_shot, data=ec_data)
summary(fit_H5)

fit_H7_vir <- aov(H7_vir_log ~ ternary_shot, data=ec_data)
summary(fit_H7_vir)

fit_H7 <- aov(H7_log ~ ternary_shot, data=ec_data)
summary(fit_H7)
```

### Comment
There is an effect of vaccinations pretty much everywhere. Exceptions are,  
for the IC50:
- H3 Brisbane (small effect),
- H4.

For the EC50 they are:
- H12,
- H3,
- H4,
- H7 virus,
- H7.

On the other hand we already knew it, since flu shots increase with age.
```{r shots_vs_age, echo=TRUE}
library(ggplot2)
library(MASS)
p <- ggplot(data=ic_data, aes(x=age, y=flushot_avg)) + geom_jitter()
p <- p + geom_smooth(method="loess")
p
```

Do titers increase with the shots alone? This is more difficult to see from a
plot only. For example, the smoothing below shows a moderate increase of H2
titer for a range of shuts between zero and ten.
```{r titers_vs_shots, echo=TRUE}
p <- ggplot(data=ic_data, aes(x=flushot_avg, H2_log)) + geom_jitter()
p <- p + geom_smooth(method="loess")
p
```

In my opinion, the correct way to discern the effect of age and vaccinations is
to perform a regression with both terms and compare it to a regression with the
age only. We do it for H2, IC50 data.

```{r age_and_shots, echo=TRUE}
fit_1 <- glm(H2_log ~ age, data=ic_data)
summary(fit_1)

fit_0 <- glm(H2_log ~ age + flushot_avg, data=ic_data)
summary(fit_0)
```
Even the AIC (Akaike's Information Criterion) is higher for the model with the
vaccinations (the lower AIC, the better).
The coefficient for the shots is very close to zero, and can be removed without
worries.

### Regression for age categories, gender and episodes
```{r regress_all}
 fit_all <- glm(H2_log ~ episodes_avg + sex + age, data=ic_data)
summary(fit_all)
```



<a id="done_stata"></a>
## 