```{r setup, include=FALSE}
# set global chunk options
library(knitr)
opts_chunk$set(cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE)
# import funtions
source("functions.R")
load("ic_ec.Rdata")
```

## Table of contents
* [IC50](#ic50)

  [H1 Brisbane](#ic50h1brisbane) [H1 PR8](#ic50h1pr8)
  
  [H2](#ic50h2)
  
  [H3 Brisbane](#ic50h3brisbane) [H3 Hong Kong](#ic50h3hk)
  
  [H5 percent at 1st dilution](#ic50h5_dil) [H7 percent at 1st dilution](#ic50h7_dil)
  
* [EC50](#ec50)
  
  [H1 pandemic 2009](#ec50h1pdm09) [H2](#ec50h2) [H3](#ec50h3)
  
  [H4](#ec50h4) [H5](#ec50h5) [H7](#ec50h7) [H7 virus](#ec50h7vir)
  
  [H12](#ec50h12)
  
----

Effect of seniority and vaccinations
------------------------------------
Here below we will study the combined effect of vaccinations and seniority on
antibody titers. We will perform analysis of covariance (ANCOVA) to discern the
separate and combined effect of the two regressors. The vaccination status is
treated as a categorical binary variable (vaccinated/non vaccinated), as the
self-reported number of vaccinations cannot be considered a reliable estimate,
while the age is a quantitative covariate.

We will perform ANCOVA estimates with the relation `shot*age`, meaning that the
model will try to estimate the antibody titer as a function of the vaccination
status, the age and their product. In symbols
$$
antibody = intercept + \alpha \cdot age + \beta \cdot shot + \gamma \cdot age \cdot shot
$$
which can be rewritten as
$$
antibody = intercept  + \beta \cdot shot + (\gamma \cdot shot + \alpha) \cdot age.
$$
If the $\gamma$ coefficient is significantly different from zero, it means that
the slope of the regression lines $(\gamma\cdot shot + \alpha)$ is different for
different categories of the vaccination status. We will perform the analysis by
calling `glm` which performs robust linear regression accounting for non
constant variance (a.k.a. heteroskedasticity).

A slow-paced discussion of ANCOVA can be found in this [blog post][1].

----

<a id="ic50"></a>
### IC50
<a id="ic50h1brisbane"></a>
#### H1 Brisbane: shot only
For H1 Brisbane, we remove first the interaction term and then also the age,
concluding that only the vaccinations are contributing to the antibody titer.
```{r ic_slopes_H1_brisbane}
fit_1 <- glm(H1_brisbane_log ~ age * shot, data=ic_data)
summary(fit_1)

fit_2 <- glm(H1_brisbane_log ~ age + shot, data=ic_data)
summary(fit_2)

fit_3 <- glm(H1_brisbane_log ~ shot, data=ic_data)
summary(fit_3)
```

<a id="ic50h1pr8"></a>
#### H1 PR8: shot only
```{r ic_slopes_h1pr8}
fit_1 <- glm(H1_PR8_log ~ age * shot, data=ic_data)
summary(fit_1)

fit_2 <- glm(H1_PR8_log ~ age + shot, data=ic_data)
summary(fit_2)

fit_3 <- glm(H1_PR8_log ~ shot, data=ic_data)
summary(fit_3)
```

<a id="ic50h2"></a>
#### H2: interaction present
Although the interaction term is not very strong, removing it doesn't improve
the fit, at least in terms of AIC. Removing the shot entirely leaves us with
an even higher AIC.
```{r ic_slopes_h2}
fit_1 <- glm(H2_log ~ age * shot, data=ic_data)
summary(fit_1)

fit_2 <- glm(H2_log ~ age + shot, data=ic_data)
summary(fit_2)
```

<a id="ic50h3brisbane"></a>
#### H3 Brisbane: shot only
```{r ic_slopes_h3_brisbane}
fit_1 <- glm(H3_brisbane_log ~ age * shot, data=ic_data)
summary(fit_1)

fit_2 <- glm(H3_brisbane_log ~ age + shot, data=ic_data)
summary(fit_2)

fit_3 <- glm(H3_brisbane_log ~ shot, data=ic_data)
summary(fit_3)
```

<a id="ic50h3hk"></a>
#### H3 Hong Kong: age only
```{r ic_slopes_h3_hk}
fit_1 <- glm(H3_HK_log ~ age * shot, data=ic_data)
summary(fit_1)

fit_2 <- glm(H3_HK_log ~ age + shot, data=ic_data)
summary(fit_2)

fit_3 <- glm(H3_HK_log ~ age, data=ic_data)
summary(fit_3)
```

<a id="ic50h4"></a>
#### H4: neither shot nor age
```{r ic_slopes_h4}
fit_1 <- glm(H4_log ~ age * shot, data=ic_data)
summary(fit_1)

fit_1 <- glm(H4_log ~ age + shot, data=ic_data)
summary(fit_1)

fit_1 <- glm(H4_log ~ shot, data=ic_data)
summary(fit_1)

```

<a id="ic50h5_dil"></a>
#### H5 effect at first dilution: shot only
```{r ic_slopes_h5_dil}
fit_1 <- glm(H5_1st_dil_percent ~ age * shot, data=ic_data)
summary(fit_1)

fit_1 <- glm(H5_1st_dil_percent ~ age + shot, data=ic_data)
summary(fit_1)

fit_1 <- glm(H5_1st_dil_percent ~ shot, data=ic_data)
summary(fit_1)
```

<a id="ic50h7_dil"></a>
#### H7 effect at first dilution: neither age nor shot
```{r ic_slopes_h7_dil}
fit_1 <- glm(H7_1st_dil_percent ~ age * shot, data=ic_data)
summary(fit_1)

fit_1 <- glm(H7_1st_dil_percent ~ age + shot, data=ic_data)
summary(fit_1)

fit_1 <- glm(H7_1st_dil_percent ~ age, data=ic_data)
summary(fit_1)
```

<a id="ec50"></a>
### EC50

<a id="ec50h1pdm09"></a>
#### H1 pandemic 2009: neither shot nor age
```{r ec_slopes_h1pdm09}
fit_1 <- glm(H1_pdm09_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H4_log ~ age + shot, data=ic_data)
summary(fit_1)

fit_1 <- glm(H4_log ~ shot, data=ic_data)
summary(fit_1)
```

<a id="ec50h1pr8"></a>
#### H1 PR8: effect of both, but no interaction
```{r ec_slopes_H1_PR8}
fit_1 <- glm(H1_PR8_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H1_pdm09_log ~ age + shot, data=ec_data)
summary(fit_1)
```

<a id="ec50h2"></a>
#### H2: small signal from the interaction, but the age only fits almost
equally well
```{r ec_slopes_H2}
fit_1 <- glm(H2_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H2_log ~ age + shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H2_log ~ age, data=ec_data)
summary(fit_1)
```

<a id="ec50h3"></a>
#### H3: both contribute, but no interaction
```{r ec_slopes_H3}
fit_1 <- glm(H3_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H3_log ~ age + shot, data=ec_data)
summary(fit_1)
```

<a id="ec50h4"></a>
#### H4: neither age nor shot
```{r ec_slopes_H4}
fit_1 <- glm(H4_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H4_log ~ age + shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H4_log ~ shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H4_log ~ age, data=ec_data)
summary(fit_1)
```

<a id="ec50h5"></a>
#### H5: both contribute, but no interaction
```{r ec_slopes_H5}
fit_1 <- glm(H5_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H5_log ~ age + shot, data=ec_data)
summary(fit_1)
```

<a id="ec50h7"></a>
#### H7: barely an effect of age
```{r ec_slopes_H7}
fit_1 <- glm(H7_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H7_log ~ age + shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H7_log ~ age, data=ec_data)
summary(fit_1)
```

<a id="ec50h7_vir"></a>
#### H7 virus: no FX
```{r ec_slopes_H7_vir}
fit_1 <- glm(H7_vir_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H7_vir_log ~ age + shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H7_vir_log ~ shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H7_vir_log ~ age, data=ec_data)
summary(fit_1)
```

<a id="ec50h12"></a>
#### H12: barely an effect of age and shot, but no interaction
```{r ec_slopes_H12}
fit_1 <- glm(H12_log ~ age * shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H12_log ~ age + shot, data=ec_data)
summary(fit_1)

fit_1 <- glm(H12_log ~ age, data=ec_data)
summary(fit_1)

fit_1 <- glm(H12_log ~ shot, data=ec_data)
summary(fit_1)
```


[1]: http://r-eco-evo.blogspot.ch/2011/08/comparing-two-regression-slopes-by.html
