```{r setup, include=FALSE}
# set global chunk options
library(knitr)
opts_chunk$set(cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE)
# import funtions
source("functions.R")
```

Analysis of antibody titers from a bunch of donors
========================================================

## Table of contents

* [Data](#data)
* **Antibody titer vs. age**

    [Linear regression](#rlm_fit)
    
    [Segmented regression](#seg_fit)
    
    [Smoothing](#smooth_fit)
* [Prediction](#pred)
* [Test H2 vaccinated vs. non vaccinated](#test_h2)

----

<a id="data"></a>
## Data
Data were loaded from several csv files (one per strain) and merged into two
dataframes, one for IC50 values and one for the EC50 values.
```{r load_data, echo=FALSE, warning=FALSE}
library(plyr)
library(reshape)
# EC data
filenames <- list.files(path="../../data/data_to_merge_EC50/", full.names=TRUE)
import.list <- llply(filenames, read.csv, sep=";")
ec_data <- merge_recurse(import.list, by="donorID")
# IC data
filenames <- list.files(path="../../data/data_to_merge_IC50/", full.names=TRUE)
import.list <- llply(filenames, read.csv, sep=";")
ic_data <- merge_recurse(import.list, by="donorID")
```

## Data cleaning
IC_50 could not be computed for H5 and H7. For these two strains, rather than
the inhibitory concentrations, the inhibition value at the first dilution is
reported, expressed in percentage.

EC50 data present two outliers (one in H2 and one in H4 titers): their value
for effective concentration was lower than -40 (in log scale) and for this
reason were removed.
```{r clean_data, echo=TRUE}
ec_data$H2_log[ec_data$H2_log < -40] <- NA
ec_data$H4_log[ec_data$H4_log < -40] <- NA
```

The following plots show the distribution of values, one boxplot per strain.
IC50 for H5 and H7 are shown separately because the scale is different
(percentage).

#### IC50
```{r ic_boxplots}
library(ggplot2)
library(reshape)
# boxplots on log scales
m <- melt(ic_data[, -c(1, 8, 9)])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
# boxplots of percentages
m <- melt(ic_data[, -c(1:7)])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
```
#### EC50
```{r ec_boxplots}
# boxplots
m <- melt(ec_data[, -1])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
```

### Merge with donors data
Donors data were taken from the file `donors_view.csv` in the `data/` directory.
We excluded donors enrolled in the Novartis study, HIV positive and those for
which the number of vaccinations is not defined.

```{r merge_donors}
donors_view <- read.csv("../../data/donors_view.csv", sep=";", quote="\"",
                          header=TRUE, na.strings=c("NULL"))
# excludes those enrolled in novartis study or hiv positive or
# for which flushot is not defined
# drop phone, email, shcs
donors_view <- subset(donors_view, novartis_study == 'n' &
  donors_view$hiv == 'n' & !is.na(flushot_avg), select=-c(phone, email, shcs))

# capitalize sex
donors_view$sex <- toupper(donors_view$sex)

# define a binary variable for vaccinated
donors_view$shot[donors_view$flushot_avg > 0] <- 'Vaccinated'
donors_view$shot[donors_view$flushot_avg == 0] <- 'Non vaccinated'

ic_data <- merge(ic_data, donors_view, by="donorID")
ec_data <- merge(ec_data, donors_view, by="donorID")
```
We are left with 271 donors, 141 females and 130 males, for which we can plot
the following characteristics.
```{r age_dist}
p <- ggplot(ec_data, aes(x=age, fill=sex)) +
  geom_histogram(binwidth=5, alpha=0.9, position="dodge") +
  ggtitle("Age distribution per gender")
p
```

```{r vaccine_sex}
p <- ggplot(ec_data, aes(x=sex, fill=shot)) + geom_histogram()
  + ggtitle("Vaccinated and non vaccinated per gender")
p
table(ec_data$shot, ec_data$sex)
```
We show the boxplots again, limited to donors considered here.
#### IC50
```{r ic_boxplots_again}
# boxplots on log scales
m <- melt(ic_data[, 2:7])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
# boxplots of percentages
m <- melt(ic_data[, 8:9])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
```
#### EC50
```{r ec_boxplots_again}
# boxplots
m <- melt(ec_data[, 2:10])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
```

We observe a few strain with outliers: H1 pandemic 2009, H2 and H4. We remove
the lowest observation on each strain.

```{r remove_outliers_ec, echo=TRUE}
ec_data$H1_pdm09_log[ec_data$H1_pdm09_log < 0] <- NA
ec_data$H2_log[ec_data$H2_log < -3] <- NA
ec_data$H4_log[ec_data$H4_log < -1.5] <- NA
```


----
<a id="rlm_fit"></a>
## Analysis of antibody titer: robust linear regression
In the following, we fit a robust linear model to the relation:
antibody-titer vs. age. Lots of plots.

### IC50

```{r plot_ic50, dpi=96, message=FALSE}
plot_antibody(ic_data, "H1_brisbane_log")
plot_antibody(ic_data, "H1_PR8_log")
plot_antibody(ic_data, "H2_log")
plot_antibody(ic_data, "H3_brisbane_log")
plot_antibody(ic_data, "H3_HK_log")
plot_antibody(ic_data, "H4_log")
plot_antibody(ic_data, "H5_1st_dil_percent")
plot_antibody(ic_data, "H7_1st_dil_percent")
```

----

### EC50
```{r plot_ec50, dpi=96, message=FALSE}
plot_antibody(ec_data, "H1_pdm09_log")
plot_antibody(ec_data, "H1_PR8_log")
plot_antibody(ec_data, "H12_log")
plot_antibody(ec_data, "H2_log")
plot_antibody(ec_data, "H3_log")
plot_antibody(ec_data, "H4_log")
plot_antibody(ec_data, "H5_log")
plot_antibody(ec_data, "H7_vir_log")
plot_antibody(ec_data, "H7_log")
```

----

<a id="seg_fit"></a>
## Specific analysis of some strains: segmented regression
In the following, we test if there is support for a breakpoint and,
consequently, we fit a segmented/non segmented generalized linear model.
For each strain, split in vaccinated and non vaccinated, we report the result
of the hypothesis test, the plot and a summary of the fit.

### IC50

#### H1 Brisbane
```{r ic_segm_h1_brisbane, message=TRUE}
segmented_analysis(ic_data, "H1_brisbane_log", "Non vaccinated")
segmented_analysis(ic_data, "H1_brisbane_log", "Vaccinated")
```

#### H1 PR8
```{r ic_segm_h1_pr8, message=TRUE}
segmented_analysis(ic_data, "H1_PR8_log", "Non vaccinated")
segmented_analysis(ic_data, "H1_PR8_log", "Vaccinated")
```

#### H2
```{r ic_segm_h2, message=TRUE}
segmented_analysis(ic_data, "H2_log", "Non vaccinated")
segmented_analysis(ic_data, "H2_log", "Vaccinated")
```

#### H3 Brisbane
```{r ic_segm_h3_brisbane, message=TRUE}
segmented_analysis(ic_data, "H3_brisbane_log", "Non vaccinated")
segmented_analysis(ic_data, "H3_brisbane_log", "Vaccinated")
```

#### H3 Hong Kong
```{r ic_segm_h3_hk, message=TRUE}
segmented_analysis(ic_data, "H3_HK_log", "Non vaccinated")
segmented_analysis(ic_data, "H3_HK_log", "Vaccinated")
```


### EC50

#### H1 pandemic 2009
```{r ec_segm_h1_pdm09, message=TRUE}
segmented_analysis(ec_data, "H1_pdm09_log", "Non vaccinated")
segmented_analysis(ec_data, "H1_pdm09_log", "Vaccinated")
```

#### H1 PR8
```{r ec_segm_h1_pr8, message=TRUE}
segmented_analysis(ec_data, "H1_PR8_log", "Non vaccinated")
segmented_analysis(ec_data, "H1_PR8_log", "Vaccinated")
```

#### H2
```{r ec_segm_h2, message=TRUE}
segmented_analysis(ec_data, "H2_log", "Non vaccinated")
segmented_analysis(ec_data, "H2_log", "Vaccinated")
```

#### H3
```{r ec_segm_h3, message=TRUE}
segmented_analysis(ec_data, "H3_log", "Non vaccinated")
segmented_analysis(ec_data, "H3_log", "Vaccinated")
```

### Specific segmented regression of IC50 for H2
On IC50 for H2, the segmented regression gives a first segment that is very
close to being horizontal. It might make sense to constrain its slope to zero
to see if we gain power in estimating less coefficients.
```{r constr_segm, echo=TRUE}
fit_1 <- glm(H2_log ~ age, family=gaussian, subset=(shot=='Non vaccinated'),
             data=ic_data)
fit_seg_1 <- segmented(fit_1, seg.Z=~age, psi=list(age=33),
                     control=seg.control(stop.if.error=TRUE))
fit_1 <- update(fit_1, .~., -age)
fit_seg_2 <- update(fit_seg_1)
```

----

<a id="smooth_fit"></a>
## Specific analysis of some strains: smoothing
Since the segmented regression gives very erratic results, we try a
non-parametric regression (_i.e._ we perform smoothing).

### IC50
```{r smooth_ic, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
for (st in names(ic_data)[2:6]){
  plot_cv_smoothed(ic_data, st, "Non vaccinated")
  plot_cv_smoothed(ic_data, st, "Vaccinated")
}
```

### EC50
```{r smooth_ec, echo=FALSE, warning=FALSE}
for (st in c("H1_pdm09_log", "H1_PR8_log", "H2_log", "H3_log")){
  plot_cv_smoothed(ec_data, st, "Non vaccinated")
  plot_cv_smoothed(ec_data, st, "Vaccinated")
}
```


<a id="pred"></a>
## Prediction
First of all, it might be instructive to look at a matrix scatterplot of all
data. Here below they plots, separated by group just to reduce cluttering.
### IC50
```{r pairs_ic}
library(car)
scatterplotMatrix( ~ H1_brisbane_log + H1_PR8_log + H2_log,
                   data=ic_data, smoother.args=list(span=0.5), main="Group 1")

scatterplotMatrix( ~ H3_brisbane_log + H3_HK_log + H4_log,
                   data=ic_data, smoother.args=list(span=0.5), main="Group 2")

```

### EC50
```{r pairs_ec}
scatterplotMatrix( ~ H1_PR8_log + H1_pdm09_log + H2_log + H12_log + H5_log,
                   data=ec_data, smoother.args=list(span=0.5), main="Group 1")

scatterplotMatrix( ~ H3_log + H4_log + H7_log + H7_vir_log,
                   data=ec_data, smoother.args=list(span=0.5), main="Group 2")

```
<a id="test_h2"></a>
## Test on EC50, H2 strain

If we perform a t-test on effective concentration for H2, we do observe a
difference between vaccinated and non vaccinated.

```{r test_h2, echo=TRUE}
t.test(H2_log ~ shot, data=ec_data)
```

On the other hand, the difference is all explained by the fact that we are
not controlling for the age. In fact, if we perform ANCOVA, age is significant,
while age is not.

```{r aov_h2, echo=TRUE}
fit_aov <- aov(H2_log ~ age + factor(shot), data=ec_data)
summary(fit_aov)
```
This is because the average age in the vaccinated is higher (42 vs. 35).
```{r test_age, echo=TRUE}
t.test(age ~ shot, data=ec_data)
```

Let's make less assumption: we fit with `glm` (that allows for non-constant
variance), and we include an interaction term.

```{r glm_h2, echo=TRUE}
glm_fit <- glm(H2_log ~ age * factor(shot), data=ec_data)
summary(glm_fit)
```
The situation does not exactly get better: we have all terms significant
(albeit not by much), but the shot status coefficient alone is negative. As a
consequence, the vaccination gives a positive contribution to the EC50 only
for donors older than 36. Overall, we don't observe a clear and univoque
contribution from the vaccination, while the effect of the age is clear.

----

## _Latest update: 26 March 2013_.
