```{r setup, include=FALSE}
# set global chunk options
library(knitr)
opts_chunk$set(cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE)
# import funtions
source("functions.R")
```

Analysis of antibody titers from a bunch of donors
========================================================
## Data
Data were loaded from several csv files (one per strain) and merged into two
dataframes, one for IC50 values and one for the EC50 values.
```{r load_data, echo=FALSE, warning=FALSE}
library(plyr)
library(reshape)
# EC data
filenames <- list.files(path="../../data/data_to_merge_EC50/", full.names=TRUE)
import.list <- llply(filenames, read.csv, sep=";")
ec_data <- merge_recurse(import.list, by="donorID")
# IC data
filenames <- list.files(path="../../data/data_to_merge_IC50/", full.names=TRUE)
import.list <- llply(filenames, read.csv, sep=";")
ic_data <- merge_recurse(import.list, by="donorID")
```

## Data cleaning
IC_50 could not be computed for H5 and H7. For these two strains, rather than
the inhibitory concentrations, the inhibition value at the first dilution is
reported, expressed in percentage.

EC50 data present two outliers (one in H2 and one in H4 titers): their value
for effective concentration was lower than -40 (in log scale) and for this
reason were removed.
```{r clean_data, echo=TRUE}
ec_data$H2_log[ec_data$H2_log < -40] <- NA
ec_data$H4_log[ec_data$H4_log < -40] <- NA
```

The following plots show the distribution of values, one boxplot per strain.
IC50 for H5 and H7 are shown separately because the scale is different
(percentage).

#### IC50
```{r ic_boxplots}
library(ggplot2)
library(reshape)
# boxplots on log scales
m <- melt(ic_data[, -c(1, 8, 9)])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
# boxplots of percentages
m <- melt(ic_data[, -c(1:7)])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
```
#### EC50
```{r ec_boxplots}
# boxplots
m <- melt(ec_data[, -1])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
```

### Merge with donors data
Donors data were taken from the file `donors_view.csv` in the `data/` directory.
We excluded donors enrolled in the Novartis study, HIV positive and those for
which the number of vaccinations is not defined.

```{r merge_donors}
donors_view <- read.csv("../../data/donors_view.csv", sep=";", quote="\"",
                          header=TRUE, na.strings=c("NULL"))
# excludes those enrolled in novartis study or hiv positive or
# for which flushot is not defined
# drop phone, email, shcs
donors_view <- subset(donors_view, novartis_study == 'n' &
  donors_view$hiv == 'n' & !is.na(flushot_avg), select=-c(phone, email, shcs))

# capitalize sex
donors_view$sex <- toupper(donors_view$sex)

# define a binary variable for vaccinated
donors_view$shot[donors_view$flushot_avg > 0] <- 'Vaccinated'
donors_view$shot[donors_view$flushot_avg == 0] <- 'Non vaccinated'

ic_data <- merge(ic_data, donors_view, by="donorID")
ec_data <- merge(ec_data, donors_view, by="donorID")
```
We are left with 271 donors, 141 females and 130 males, for which we can plot
the following characteristics.
```{r age_dist}
p <- ggplot(ec_data, aes(x=age, fill=sex)) +
  geom_histogram(binwidth=5, alpha=0.9, position="dodge") +
  ggtitle("Age distribution per gender")
p
```

```{r vaccine_sex}
p <- ggplot(ec_data, aes(x=sex, fill=shot)) + geom_histogram()
  + ggtitle("Vaccinated and non vaccinated per gender")
p
table(ec_data$shot, ec_data$sex)
```
We show the boxplots again, limited to donors considered here.
#### IC50
```{r ic_boxplots_again}
# boxplots on log scales
m <- melt(ic_data[, 2:7])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
# boxplots of percentages
m <- melt(ic_data[, 8:9])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
```
#### EC50
```{r ec_boxplots_again}
# boxplots
m <- melt(ec_data[, 2:10])
p <- ggplot(m, aes(x=variable, y=value)) + geom_boxplot()
p
```

----
## Analysis of antibody titer
In the following, we fit a robust linear model to the relation:
antibody-titer vs. age. Lots of plots.

### IC50

```{r plot_ic50, dpi=96, message=FALSE}
plot_antibody(ic_data, "H1_brisbane_log")
plot_antibody(ic_data, "H1_PR8_log")
plot_antibody(ic_data, "H2_log")
plot_antibody(ic_data, "H3_brisbane_log")
plot_antibody(ic_data, "H3_HK_log")
plot_antibody(ic_data, "H4_log")
plot_antibody(ic_data, "H5_1st_dil_percent")
plot_antibody(ic_data, "H7_1st_dil_percent")
```

----

### EC50
```{r plot_ec50, dpi=96, message=FALSE}
plot_antibody(ec_data, "H1_pdm09_log")
plot_antibody(ec_data, "H1_PR8_log")
plot_antibody(ec_data, "H12_log")
plot_antibody(ec_data, "H2_log")
plot_antibody(ec_data, "H3_log")
plot_antibody(ec_data, "H4_log")
plot_antibody(ec_data, "H5_log")
plot_antibody(ec_data, "H7_vir_log")
plot_antibody(ec_data, "H7_log")
```

----

## Specific analysis of some strains
In the following, we test if there is support for a breakpoint and,
consequently, we fit a segmented/non segmented generalized linear model.
For each strain, split in vaccinated and non vaccinated, we report the result
of the hypothesis test, the plot and a summary of the fit.

### IC50

#### H1 Brisbane
```{r ic_segm_h1_brisbane, message=TRUE}
segmented_analysis(ic_data, "H1_brisbane_log", "Non vaccinated")
segmented_analysis(ic_data, "H1_brisbane_log", "Vaccinated")
```

#### H1 PR8
```{r ic_segm_h1_pr8, message=TRUE}
segmented_analysis(ic_data, "H1_PR8_log", "Non vaccinated")
segmented_analysis(ic_data, "H1_PR8_log", "Vaccinated")
```

#### H2
```{r ic_segm_h2, message=TRUE}
segmented_analysis(ic_data, "H2_log", "Non vaccinated")
segmented_analysis(ic_data, "H2_log", "Vaccinated")
```

#### H3 Brisbane
```{r ic_segm_h3_brisbane, message=TRUE}
segmented_analysis(ic_data, "H3_brisbane_log", "Non vaccinated")
segmented_analysis(ic_data, "H3_brisbane_log", "Vaccinated")
```

#### H3 Hong Kong
```{r ic_segm_h3_hk, message=TRUE}
segmented_analysis(ic_data, "H3_HK_log", "Non vaccinated")
segmented_analysis(ic_data, "H3_HK_log", "Vaccinated")
```


### EC50

#### H1 Brisbane
```{r ec_segm_h1_pdm09, message=TRUE}
segmented_analysis(ec_data, "H1_pdm09_log", "Non vaccinated")
segmented_analysis(ec_data, "H1_pdm09_log", "Vaccinated")
```

#### H1 PR8
```{r ec_segm_h1_pr8, message=TRUE}
segmented_analysis(ec_data, "H1_PR8_log", "Non vaccinated")
segmented_analysis(ec_data, "H1_PR8_log", "Vaccinated")
```

#### H2
```{r ec_segm_h2, message=TRUE}
segmented_analysis(ec_data, "H2_log", "Non vaccinated")
segmented_analysis(ec_data, "H2_log", "Vaccinated")
```

#### H3 Brisbane
```{r ec_segm_h3_, message=TRUE}
segmented_analysis(ec_data, "H3_log", "Non vaccinated")
segmented_analysis(ec_data, "H3_log", "Vaccinated")
```

## For now, analysis ending here. _19 March 2013_.

----

### H3
The results on the EC50 show that there is an effect of both the age and the
vaccination status.
```{r h3_ec, echo=TRUE}
ec_fit <- aov(H3_log ~ age + shot, data=ec_data)
summary(ec_fit)
```

On the other hand, the IC50 data only show the effect of the age, without a
difference between vaccinated and non vaccinated.
```{r h3_ic, echo=TRUE}
ic_fit <- aov(H3_HK_log ~ age + shot, data=ic_data)
summary(ic_fit)
```
